Quantifying the importance of an inference model in Bayesian phylogenetics

### 2019-10-07 content.tex
Processed feedback @rsetienne, fix https://github.com/richelbilderbeek/razzo/issues/313

The second step (Fig.~\ref{fig:pipeline}, 3)
selects one or more inference models $I$ from a set of inference 
models $I_{1},\dots,I_{n}$. 
We define an experiment $X_{i}$ as the combination of 
an inference model $I_{i}$ and the conditions $C_{i}$ 
to actually use it in the inference step.
For example, we may require that an inference
model (a combination of a tree prior, clock model and site model) 
should include the generative/true tree prior. 
As a second example, we may require that we have selected a set of 
candidate inference models,
of which only the best should be used in the actual inference.
In the first example, we specified the condition $C_{i}$ that this
generative model should always be run, whereas in the second example,
we specified condition $C_{i}$ that a candidate model should only be run
when it is the best.
The 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood), given the alignment 
simulated in the previous step.
The evidence for an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package. 
We note that scripted use of BEAST2 packages is only possible under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences 
using the web interface of \verb;mcbette; [\cite{mcbette}].

@@

This can be seen for both the "generative" and "candidate" model 
types (see Table~\ref{tab:options}).
If the chosen model type is "generative", 
the tree prior chosen for the twin inference will be 
exactly the same used to generate the tree.
In the main pipeline, as the tree model $p_{G}$ is non-standard, 
it cannot be used in inference.
If, instead, the chosen model type is "candidate", 
the twin tree prior will be included in the pool of examined models 
during the process of selection of the inference model.

Finally, if the goal is to evaluate BEAST2's performance on a non-standard tree prior, 
the last source of stochasticity comes from phylogenies. 
In fact, a single phylogeny cannot be considered as fully representative of the prior. 
For this reason multiple phylogenies, 
as well as an equal number of twins, 
must be considered. 
Having the error measure normalized (i.e. comprised in the interval $[0, 1]$), 
it is possible to considerate the aggregated versions of the errors distributions 
across all the runs.
Therefore, if the number of considered phylogenies is high enough, 
the comparison between the main pipeline's aggregated error distribution 
and its twin counterpart leads to a fair evaluation 
of the new tree prior with respect to the baseline error.

@@

In this example we use a standard generative tree model $\mathit{p_{G}^0}$, 
namely the Yule (pure-birth) tree prior. 
We choose to use a small tree with six taxa, to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but it ties in with the mutation rate 
used in simulating an alignment in the next step.

@@

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$ is by (our) definition
non-standard, and hence not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is which standard tree prior should be used in 
the inference. In this example, we do not only re-use our hand-picked 
inference model, but we also pick the best inference model from a set of 
inference models, i.e.  the model with 
the highest evidence measured by its marginal likelihood. 
We show the procedure using \verb;pirouette; to answer a second 
research question: "What is the inference error made on a novel phylogeny 
when using the best inference model, in comparison to a hand-picked model?

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_3}. 
While the error distributions using the best or generative model 
as inference model are very similar, 
the error distributions of the true tree are substantially 
larger than those of the twin tree. 
This is the error made in inference 
by not using the generating species tree prior.

>>>>

The second step (Fig.~\ref{fig:pipeline}, 3)
selects one or more inference models $I$ from a set of inference 
models $I_{1},\dots,I_{n}$. 
We define an experiment $X_{i}$ as the combination of 
an inference model $I_{i}$ and the conditions $C_{i}$ 
to actually use it in the inference step.
For example, we may require that an inference
model (a combination of a tree model, clock model and site model) 
should include the generative/true tree model. 
As a second example, we may require that we have selected a set of 
candidate inference models,
of which only the best should be used in the actual inference.
In the first example, we specified the condition $C_{i}$ that this
generative model should always be run, whereas in the second example,
we specified condition $C_{i}$ that a candidate model should only be run
when it is the best.
The 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood), given the alignment 
simulated in the previous step.
The evidence for an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package. 
We note that scripted use of BEAST2 packages is only possible under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences 
using the web interface of \verb;mcbette; [\cite{mcbette}].

@@

This can be seen for both the "generative" and "candidate" model 
types (see Table~\ref{tab:options}).
If the chosen model type is "generative", 
the tree prior chosen for the twin inference will  
exactly match the model to generate the tree.
In the main pipeline, as the tree model $p_{G}$ is non-standard, 
it cannot be used in inference.
If, instead, the chosen model type is "candidate", 
the twin tree model will be included in the pool of examined models 
during the process of selection of the inference model.

Finally, if the goal is to evaluate BEAST2's performance on a non-standard tree prior, 
the last source of stochasticity comes from phylogenies. 
In fact, a single phylogeny cannot be considered 
as fully representative of the model. 
For this reason multiple phylogenies, 
as well as an equal number of twins, 
must be considered. 
Having the error measure normalized (i.e. comprised in the interval $[0, 1]$), 
it is possible to considerate the aggregated versions of the errors distributions 
across all the runs.
Therefore, if the number of considered phylogenies is high enough, 
the comparison between the main pipeline's aggregated error distribution 
and its twin counterpart leads to a fair evaluation 
of the new tree prior with respect to the baseline error.

@@

In this example we use a standard generative tree model $\mathit{p_{G}^0}$, 
namely the Yule (pure-birth) tree model. 
We choose to use a small tree with six taxa, to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but it ties in with the mutation rate 
used in simulating an alignment in the next step.

@@

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree model $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$ is by (our) definition
non-standard, and hence not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is which standard tree prior should be used in 
the inference. In this example, we do not only re-use our hand-picked 
inference model, but we also pick the best inference model from a set of 
inference models, i.e.  the model with 
the highest evidence measured by its marginal likelihood. 
We show the procedure using \verb;pirouette; to answer a second 
research question: "What is the inference error made on a novel phylogeny 
when using the best inference model, in comparison to a hand-picked model?

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_3}. 
While the error distributions using the best or generative model 
as inference model are very similar, 
the error distributions of the true tree are substantially 
larger than those of the twin tree. 
This is the error made by the mismatch
between the generating species tree model and the tree prior used in inference.
